Demystifying the Computer Science \\ PhD Admission in the US\\{\large A Handbook for International \emph{and Domestic} Students

### 2024-03-20 demystify.tex

\begin{enumerate}
  \item Should you apply?
        \begin{itemize}
          \item \emph{Yes, definitely}.  CS PhD study in the US is fully funded and admission into good universities is not any harder than in other countries (\autoref{sec:should}).
        \end{itemize}
  \item How is your application evaluated?
        \begin{itemize}
          \item Applications are evaluated by the \emph{PhD Admission} (adcom) committee and each application is typically reviewed by three faculty (\autoref{sec:evalapps}).
          \item Individually faculty \emph{cannot directly admit} a student---so do not email and ask if you have a chance. However, faculty can \emph{advocate} for a student and therefore increase their admission chance---contact and introduce yourself (\autoref{sec:contact}).
         \end{itemize}
  \item Application Materials
        \begin{itemize}
          \item The committee will look at various factors, but the most important ones are research ability, e.g., publications, statement of purpose (SOP), and  letters of recommendation (LORs).
          \item \emph{LORs are very important}, but only if they are personalized and talk about your research ability (\autoref{sec:lor}).
          \item \emph{SOP is very important}. Write it in such a way that makes you \emph{stand out} (\autoref{sec:research-statement} and \autoref{sec:improve-your-chance})
          \item  \emph{Pubication can greatly help}. Publications are good venues are concrete evidence that you have successfully engaged in research. But they are not required! (\autoref{sec:research-ability}).

@@

\section{Research Ability}\label{sec:research-ability}

A concrete evidence of research ability is having \textbf{papers in reputable international journals or conferences}.
Having published papers, especially at top venues, is a sign that the applicant has successfully involved in research.  Students  accepted to top schools often have multiple first-authored papers at top venues.

Additionally, \textbf{work experiences at well-known research laboratories}, such as Microsoft Research, can strengthen your
application.  Unfortunately, many good research places in your countries, e.g., VinAI in Vietnam, remain relatively unknown to most universities in the US. So you or your LoR writers should explicit say something about them in your statement.

Finally, \textbf{participating internationally recognized competitions} can also demonstrate your research potential.
For example, participating in Math Olympiads if you want to do theory or  winning ACM programming contests if you want to ``build'' systems, e.g., software analysis.

@@

@@

\section{Standard Tests (GREs, IELTS, etc) and Work Experiences}\label{sec:standard-tests}

@@

\paragraph{Work Experiences} Work experiences, especially if they are not research-based, do not carry much weight. Similarly, LoR from your supervisor for non-research experience will not count much.

\textbf{Advice}: do not spend much time talking about non-research related job in your SOP.
However, if you have research-based work experiences, then they can be useful and should be mentioned in your SOP and LORs. For example, if you have worked at a research lab (e.g., Microsoft, DoE, VinAI in Vietnam, etc) (\autoref{sec:research-ability}), then you should mention that in your SOP and ask your supervisor to write about your research experiences and potentials in their LOR (\autoref{sec:lor}).

@@

\textbf{Advice}: Prepare your CV in such away that it allows the reviewers to quickly scan to identify major achievements (e.g., Publications, Programming Competition Awards, Teaching Experience).

@@

\paragraph{If you do not get admitted} If you do not get admitted to any schools or don't want to go to the ones that admit you, try again next time.  Graduate admission can involve randomness and noise. In the meantime, you can work on improving your profile, e.g., get more research experiences, publish more papers, improve your connections for better LoR writers, etc. See rejection reasons (\autoref{sec:why-rejected}) and additional advice ()\autoref{sec:improve-your-chance}).

@@

As discussed in \autoref{sec:undergrad-research}, research experience is a crucial component of a strong PhD application. It gives you opportunities to try out research, determine what research area you're interested in, publish papers (\autoref{sec:research-ability}), and get strong LoRs (\autoref{sec:lor}).

>>>>

\begin{enumerate}
  \item Should you apply?
        \begin{itemize}
          \item \emph{Yes, definitely}.  CS PhD study in the US is fully funded and admission into good universities is not any harder than in other countries (\autoref{sec:should}).
        \end{itemize}
  \item How is your application evaluated?
        \begin{itemize}
          \item Applications are evaluated by the \emph{PhD Admission} (adcom) committee and each application is typically reviewed by three faculty (\autoref{sec:evalapps}).
          \item Individually faculty \emph{cannot directly admit} a student---so do not email and ask if you have a chance. However, faculty can \emph{advocate} for a student and therefore increase their admission chance---contact and introduce yourself (\autoref{sec:contact}).
         \end{itemize}
  \item Application Materials
        \begin{itemize}
          \item The committee will look at various factors, but the most important ones are research ability, e.g., publications, statement of purpose (SOP), and  letters of recommendation (LORs).
          \item \emph{LORs are very important}, but only if they are personalized and talk about your research ability (\autoref{sec:lor}).
          \item \emph{SOP is very important}. Write it in such a way that makes you \emph{stand out} (\autoref{sec:research-statement} and \autoref{sec:improve-your-chance})
          \item  \emph{Pubication can greatly help}. Publications are good venues are concrete evidence that you have successfully engaged in research. But they are not required! (\autoref{sec:research-experiences}).

@@

@@

\section{Research Experiences}\label{sec:research-experiences}

\paragraph{Publications} A concrete evidence of research ability is having \emph{papers in reputable international journals or conferences}.
Having published papers, especially at top venues, is a sign that the applicant has successfully involved in research.  Students  accepted to top schools often have multiple first-authored papers at top venues.

\paragraph{Research Experiences} Work experiences at \emph{well-known research laboratories}, such as Microsoft Research, can strengthen your
application.  Unfortunately, many good research places in your countries, e.g., VinAI in Vietnam, remain relatively unknown to most universities in the US. So you or your LoR writers should explicit say something about them in your statement.

\paragraph{Competitions} Winning \emph{internationally recognized competitions} can also demonstrate your research potential.
For example, participating in Math Olympiads if you want to do theory or  winning ACM programming contests if you want to ``build'' systems, e.g., software analysis.

@@

\section{Standard Tests (GREs, IELTS, etc), Work Experiences, Online Courses, and Certificates}\label{sec:standard-tests}

@@

\paragraph{Work Experiences} Work experiences, especially if they are not research-based, \emph{do not} carry much weight. Similarly, LoR from your supervisor for non-research experience will not count much.

\textbf{Advice}: do not spend much time talking about non-research related job in your SOP.
However, if you have research-based work experiences, then they can be useful and should be mentioned in your SOP and LORs. For example, if you have worked at a research lab (e.g., Microsoft, DoE, VinAI in Vietnam, etc) (\autoref{sec:research-experiences}), then you should mention that in your SOP and ask your supervisor to write about your research experiences and potentials in their LOR (\autoref{sec:lor}).

\paragraph{Online Courses and Certificates}
In general, these do not carry much weight as they do not show your research ability. We do not care much if you have taken on online Coursera AI course or have a professional certificate from Microsoft.
However, as mentioned in \autoref{sec:non-stem}, if you do not have a CS background, you can use these to show your ability to handle CS courses.

@@

\textbf{Advice}: Prepare your CV in such away that it allows the reviewers to quickly scan to identify major achievements, e.g., Publications, Programming Competition Awards, Teaching Experience.

@@

\paragraph{If you do not get admitted} If you do not get admitted to any schools or don't want to go to the ones that admit you, try again next time.  Graduate admission can involve randomness and noise. In the meantime, you can work on improving your profile, e.g., get more research experiences, publish more papers, improve your connections for better LoR writers, etc. See rejection reasons (\autoref{sec:why-rejected}) and additional advice (\autoref{sec:improve-your-chance}).

@@

As discussed in \autoref{sec:undergrad-research}, research experience is a crucial component of a strong PhD application. It gives you opportunities to try out research, determine what research area you're interested in, publish papers (\autoref{sec:research-experiences}), and get strong LoRs (\autoref{sec:lor}).
